Isolator: accurate and consistent analysis of RNA-Seq experiments

### 2015-08-05 paper.tex
More revision. Start adding junk to supplement.

\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{sectsty}
\usepackage{ctable}
\usepackage{hyperref}

%\setromanfont{Adagio_Serif Light}
%\setsansfont[BoldFont={Adagio_Sans}]{Adagio_Sans Light}
%\allsectionsfont{\normalfont\sffamily\bfseries}

@@

\begin{abstract}
    While RNA-Seq has enabled great progress towards the goal of wide-scale
    isoform-level mRNA quantification, short reads have limitations when
    resolving complex and similar sets of isoforms. As a result, estimates
    of isoform abundance, compared to those at the gene level, carry far
    more uncertainty. When confronted with this uncertainty, commonly used
    methods produce estimates that are unstable; small perturbations in the
    data often produce dramatically different results. We introduce a new
    method, Isolator, which analyzes all samples in an experiment in unison
    using a simple Bayesian hierarchical model. Combined with aggressive bias
    correction, it produces estimates that are simultaneously accurate and
    stable. In a comprehensive comparison of accuracy and stability, we show
    that this property is unique to Isolator. We further demonstrate that
    the approach of modeling an entire experiment enables new analyses by
    examining splicing monotonicity across several time points in the
    development of human cardiomyocyte cells.
\end{abstract}

@@

Since the introduction of RNA-Seq it has rapidly become a preferred
method of studying gene expression, having proved to be reliable,
reproducible, and increasingly affordable. In principle, RNA-Seq enables
measurements of expression at a higher resolution than has previously
been possible or practical at the same scale. Rather than an often
vaguely defined notion of a gene, expression of specific isoform or
exons can be measured. In practice this promise is difficult to realize.
There are no standardized names or symbols for isoforms, annotations are
often incomplete or in disagreement, and methods designed to detect
changes in splicing rarely concur (TODO: cite).

Because a short read is often compatible with many isoforms, the signal
produced from sequencing must be deconvoluted, requiring a certain level
of algorithmic sophistication. This problem was tackled early on and
with great success by Cufflinks, and indeed many of the methods
developed in the interim have followed from the same basic model of
RNA-Seq in which transcripts are represented as particular distributions
over possible reads, and the problem is then to infer mixing
coefficients (relative expression values) inducing a mixture model that
best explains the data.

@@

An efficient MCMC algorithm is used to generate samples over the
parameters of this model, which are saved in an HDF5 based format. The
output from the sampler can then be processed to produce point
estimates, credible intervals, posterior probabilities, and diagnostics.
Though designed to be run on multiple samples concurrently, Isolator can
be run as a conventional tool on individual samples.

The broad effect of this model is to encode the common assumption that
genes tend to be similarly expressed and similarly spliced between
replicates, and to a lesser extent between conditions. In the absence of
sufficient evidence to the contrary, this informative prior shrinks
estimates towards a baseline of no change, producing more conservative
estimates of effect size than one would obtain by using, for example, a
flat Dirichlet prior. This feature of the model is particularly
important for robust treatment of low-expression genes. With few reads
there is little power to determine splicing patterns. Considering the
samples in isolation will often produce very different estimates.
Statistical tests based on these estimates, without considering the full
posterior distribution, make approximations by either either
disregarding the relative or compositional nature of the data and
assuming independence, or by relying on asymptotic properties that may
not hold when a gene is not deeply sequenced.

In addition to approach to we attempt to aggressively correct for
technical effects. The naive model of RNA-Seq assumes fragments are
sampled uniformly at random from transcripts in proportion to their
abundance. Several have observed that this uniformity assumption is far
from an accurate description of real sequencing data (TODO: cite).
Though severe at the nucleotide resolution, gene level expression
estimates are only moderately affected by this bias, as it tends to
average out over long transcripts. When one considers expression at the
isoform level, attempting to correct for this nonuniformity becomes
vital, as the difference between two isoforms is often only a few
nucleotides.

We previously developed an efficient and versatile model to account for
sequence specific bias that commonly occur at the ends of fragments
(TODO: cite). In Isolator, we supplemented this with methods accounting
for 3' bias, fragment GC-content bias, and positional bias caused by
end-effects of fragmentation.

@@

\section{Results}\label{results}

The convention for evaluating the accuracy of RNA-Seq analysis methods
has consisted chiefly of comparing point estimate accuracy in
simulations, supplemented with comparisons to qPCR or other
technologies. Tuning methods exclusively to maximize accuracy on
simulated data creates the risk of solving only an idealized
mathematical model of RNA-Seq while disregarding noise, bias, and
nonuniformity. As no gold standard for whole-transcriptome isoform-level
estimation exists, we adopt multi-faceted approach, with a focus
measuring the stability or consistency of estimates, along with the
accuracy.

Comparisons of gene or transcript expression is sensitive to how one
chooses to measure similarity or distance. Because gene expression
typically varies across several orders of magnitude, measurements like
the Pearson correlation, $L_1$ or $L_2$ distance, or root-mean-square
error are often dominated by a small subset of the most highly expressed
genes. Recognizing this, a popular alternative is the Spearman rank
correlation, which conversely can overemphasize the majority of
transcripts which have no, or very low expression. These issues are
discussed in depth by Lovell et al (Lovell et al. 2015), who propose
measurements of proportionality as a more meaningful metric. We adopt
their ``proportionality correlation'', defined as
$\text{pcor}(x, y) = 2\text{cov}(\log(x), \log(y)) / (\text{var}(\log(x)) + \text{var}(\log(y)))$.
Similar to other measures of correlation, proportionality correlation
varies from -1 to 1, with 1 indicating perfectly proportionality, and -1
perfect reciprocality. Zeros are accounted for by adding 0.1 to all TPM
values.

Many methods have been proposed to aid in the analysis of RNA-Seq
experiments. We limit our discussion specifically to those that seek to
estimate the relative abundance a set of known transcripts. These
include Cufflinks, RSEM, eXpress, BitSeq, Sailfish, Salmon, BitSeq, and
Kallisto. RSEM was used to produce both maximum likelihood and posterior
mean estimates, which we label ``RSEM/ML'' and ``RSEM/PM'' respectively.
This is not an exhaustive account of such methods, but does represent a
wide variety of popular approaches.

\subsection{Agreement with qPCR and spike-in
controls}\label{agreement-with-qpcr-and-spike-in-controls}

@@

We use data generated by the SEQC project, consisting of four reference
samples (labeled A, B, C, D). The RNA-seq methods were run using
approximately 25 million reads per sample, obtaining estimates that were
then compared to qPCR in Tables TODO and TODO, and the known ERCC
spike-in mixtures in Table TODO. In each of these we find that Isolator
produces the highest correlation of the methods compared, though
sometimes by negligible margin. Despite two different technologies, and
two different sets of genes, the qPCR benchmarks agree closely, ranking
the methods nearly identically. Cufflinks, eXpress, Salmon, and RSEM/ML
typically perform similarly while BitSeq, Kallisto, Sailfish, and
RSEM/PM trail slightly. Results are similar with ERCC spike-in controls,
the notable exceptions being Kallisto and Sailfish, which show higher
accuracy, and eXpress, which shows relatively lower accuracy.

\begin{table}
\begin{tabular}[c]{@{}lrrrr@{}}
\toprule\addlinespace
Method & A & B & C & D
\\\addlinespace
\midrule
Isolator & \textbf{0.903} & \textbf{0.908} & \textbf{0.890} &
\textbf{0.884}
\\\addlinespace
eXpress & 0.901 & 0.902 & 0.884 & 0.874
\\\addlinespace
Cufflinks & 0.900 & 0.901 & 0.865 & 0.872
\\\addlinespace
Salmon & 0.899 & 0.897 & 0.882 & 0.871
\\\addlinespace
RSEM/ML & 0.899 & 0.896 & 0.881 & 0.870
\\\addlinespace
Kallisto & 0.893 & 0.885 & 0.874 & 0.860
\\\addlinespace
BitSeq & 0.875 & 0.865 & 0.863 & 0.848
\\\addlinespace
Sailfish & 0.879 & 0.857 & 0.854 & 0.834
\\\addlinespace
RSEM/PM & 0.878 & 0.866 & 0.864 & 0.848
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between gene level quantification
of 806 genes using TaqMan qPCR and RNA-Seq quantification.}
\end{table}

\begin{table}\begin{tabular}[c]{@{}lrrrr@{}}
\toprule\addlinespace
Method & A & B & C & D
\\\addlinespace
\midrule
Isolator & \textbf{0.878} & \textbf{0.866} & \textbf{0.839} &
\textbf{0.852}
\\\addlinespace
Cufflinks & 0.870 & 0.856 & 0.799 & 0.841
\\\addlinespace
eXpress & 0.870 & 0.855 & 0.829 & 0.840
\\\addlinespace
Salmon & 0.866 & 0.852 & 0.826 & 0.836
\\\addlinespace
RSEM/ML & 0.865 & 0.851 & 0.825 & 0.835
\\\addlinespace
BitSeq & 0.840 & 0.821 & 0.802 & 0.813
\\\addlinespace
Kallisto & 0.858 & 0.840 & 0.817 & 0.826
\\\addlinespace
Sailfish & 0.844 & 0.814 & 0.797 & 0.802
\\\addlinespace
RSEM/PM & 0.840 & 0.822 & 0.803 & 0.811
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between gene level quantification
of 18353 genes using PrimePCR qPCR and RNA-Seq quantification.}
\end{table}

\begin{table}\begin{tabular}[c]{@{}lrrrr@{}}
\toprule\addlinespace
Method & A & B & C & D
\\\addlinespace
\midrule
Isolator & \textbf{0.979} & \textbf{0.978} & \textbf{0.981} &
\textbf{0.982}
\\\addlinespace
Salmon & 0.976 & 0.975 & 0.978 & 0.979
\\\addlinespace
Kallisto & 0.972 & 0.972 & 0.973 & 0.976
\\\addlinespace
Sailfish & 0.970 & 0.969 & 0.969 & 0.972
\\\addlinespace
Cufflinks & 0.967 & 0.969 & 0.972 & 0.974
\\\addlinespace
RSEM/PM & 0.943 & 0.949 & 0.944 & 0.949
\\\addlinespace
RSEM/ML & 0.941 & 0.948 & 0.945 & 0.951
\\\addlinespace
BitSeq & 0.940 & 0.949 & 0.943 & 0.946
\\\addlinespace
eXpress & 0.931 & 0.939 & 0.935 & 0.942
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between known proportions of 92
ERCC spike-in controls and RNA-Seq quantification.}
\end{table}

@@

The four SEQC samples consist of two commercially available reference
RNA samples, labeled $A$ and $B$, and two synthetic samples formed by
mixing $A$ and $B$ in specific proportions. Sample $C$ is composed of
75\% $A$ and 25\% $B$, while sample $D$ is 25\% $A$ and 75\% $B$. This
design allows us to directly measure expression consistency. If we
estimate transcript abundance for each sample producing estimates
$a, b, c, d$, then for these estimates to be in agreement we should
expect $c$ to be approximated by $0.25a + 0.75b$, and similarly $d$ by
$0.75a + 0.25b$. In this manner we measured consistency for each method
using proportionality correlation (Table TODO).

@@

Estimates made at the gene level (Supplementary Table TODO) agree with
these results, but are of uniformly higher correlation and show a much
smaller gap between posterior mean and maximum likelihood estimate.
Posterior mean methods (Isolator, BitSeq, RSEM/PM) again show the
highest correlation and are in very close agreement.

\begin{table}\begin{tabular}[c]{@{}lrr@{}}
\toprule\addlinespace
Method & $c$ vs $0.75a + 0.25b$ & $d$ vs $0.25a + 0.75b$
\\\addlinespace
\midrule
Isolator & \textbf{0.975} & \textbf{0.975}
\\\addlinespace
BitSeq & 0.967 & 0.967
\\\addlinespace
RSEM/PM & 0.968 & 0.967
\\\addlinespace
Sailfish & 0.932 & 0.925
\\\addlinespace
RSEM/ML & 0.922 & 0.919
\\\addlinespace
Salmon & 0.916 & 0.914
\\\addlinespace
Kallisto & 0.907 & 0.902
\\\addlinespace
eXpress & 0.903 & 0.899
\\\addlinespace
Cufflinks & 0.870 & 0.916
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between transcript-level estimates
for the mixed samples C and D and weighted averages of estimates for A
and B, corresponding to the mixture proportions for C and D.}
\end{table}

\begin{table}\begin{tabular}[c]{@{}lrr@{}}
\toprule\addlinespace
Method & $c$ vs $0.75a + 0.25b$ & $d$ vs $0.25a + 0.75b$
\\\addlinespace
\midrule
Isolator & \textbf{0.995} & \textbf{0.995}
\\\addlinespace
BitSeq & 0.994 & 0.994
\\\addlinespace
RSEM/PM & 0.994 & 0.993
\\\addlinespace
RSEM/ML & 0.987 & 0.987
\\\addlinespace
Salmon & 0.986 & 0.986
\\\addlinespace
eXpress & 0.983 & 0.983
\\\addlinespace
Sailfish & 0.978 & 0.976
\\\addlinespace
Kallisto & 0.972 & 0.970
\\\addlinespace
Cufflinks & 0.941 & 0.982
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between gene-level estimates for
the mixed samples C and D and weighted averages of estimates for A and
B, corresponding to the mixture proportions for C and D. TODO: move this
table to the supplement.}
\end{table}

@@

\begin{figure*}
\includegraphics[width=\textwidth]{batch-effect-figure.pdf}
\caption{
\textbf{a} A heatmap showing pairwise proportionality correlation between the sample
sampled sequenced on two flowcells each at five sites. Flowcells are numbered 1
or 2 and sequencing sites are abbreviated with three letter codes: Australian
Genome Research Facility (AGR), Beijing Genome Institute (BGI), Cornell
University (CNL), Mayo Clinic (MAY), and Novartis (NVS). \textbf{b} The absolute
change in correlation induced by enabling bias correction on methods that
support it. For clarity this plot excludes points for BitSeq estimates of "MAY
2", as bias correction has an extreme detrimental effect on these.
}
\end{figure*}

We further examined the question of consistency by comparing the same
samples sequenced on different flow cells and at different sites, again
using data from the SEQC. We used 10 samples from the larger SEQC
dataset, each constituting a single flowcell lane from two separate
flowcells sequenced at five different sites, all on Illumina HiSeq 2000
instruments, and compared pairwise agreement (Figure TODO).

These correlations are somewhat smaller than those in Table TODO,
largely because of shallower sequencing: there five flowcell lanes were
combined, totaling \textasciitilde{}25 millions reads per sample, here
only one lane or \textasciitilde{}5 million reads are used.
Nevertheless, we again posterior mean methods with showing much higher
agreement between pairs.

When accounting for batch effects, bias correction methods can have a
significant impact. To examine the efficacy of bias correction, we
disabled bias correction functionality on those methods that support it,
repeated the experiment, and measured the change in pairwise correlation
(Figure TODO). We see that bias correction is largely beneficial for
each method,

Isolator, Salmon, and Cufflinks show similar improvements with bias
correction, though in a few cases Cufflinks slightly decreases
agreement. Kallisto show a consistent, but slight improvement. BitSeq's
bias correction sometimes has a very positive effect, but other times
negative or disastrous effect. Estimates from the ``MAY 2'' samples in
particular had far worse agreement with other samples with bias
correction enabled.

While this benchmark is informative, it should be considered a lower
bound on the batch effects and bias: these samples were identically
prepared and sequenced. The fact that technical effects is indicative
that RNA-Seq is not so consistent that batch effects can be under any
circumstanced ignored.

@@

To demonstrate Isolator's capacity to produce more accurate estimates by
modeling entire experiments, we simulated a simple RNA-Seq experiment
consisting of two conditions each with three replicated. Expression
values were generated from a two-component log-normal model with
parameters fit to Cufflinks estimates of data from Kuppusamy et al
(TODO: cite). We generated simulated RNA-Seq reads using rlsim (TODO:
cite).

In this benchmark, Isolator significantly improves on existing methods.
It has the unique advantage of sharing information between replicates.
With the assumption that genes tend to be similarly spliced between
replicates, we are able to more effectively resolve transcript
expression in complex loci. Kallisto, Salmon, RSEM/ML, and Cufflinks all
perform very similarly. While comparisons to qPCR showed eXpress and
Cufflinks with a significant advantage, effective bias correction is
less important in simulated data, plausibly explaining the difference.
Although rlsim models some forms of bias, technical effects in RNA-Seq
are not entirely understood, so these effects are likely understated.

\begin{table}\begin{tabular}[c]{@{}ll@{}}
\toprule\addlinespace
Method & Correlation
\\\addlinespace
\midrule
Isolator & 0.919
\\\addlinespace
Kallisto & 0.887
\\\addlinespace
Salmon & 0.886
\\\addlinespace
RSEM/ML & 0.881
\\\addlinespace
Cufflinks & 0.881
\\\addlinespace
eXpress & 0.825
\\\addlinespace
Sailfish & 0.816
\\\addlinespace
RSEM/PM & 0.806
\\\addlinespace
BitSeq & 0.796
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between ground truth and estimates
produced by each method on simulated RNA-Seq.}
\end{table}

@@

RNA-Seq reads become more informative with length and quantity. We
compared data from the same sample sequenced twice, once with 300pb
paired-end reads on a MiSeq sequencer and again with 100bp paired-end
reads using a HiSeq 2000, yielding approximately 4.5 million and 11
million reads, respectively. To elucidate the effect of deeper
sequencing, we sampled without replacement smaller subsets of the 11
million HiSeq 2000 reads and measured correlation between estimates
these subsamples with estimates from the MiSeq data (Figure TODO).

>>>>

\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{sectsty}
\usepackage{ctable}
\usepackage{hyperref}
\usepackage[numbers]{natbib}
\usepackage{xr}

\externaldocument{supplement}

\setromanfont{Adagio_Serif Light}
\setsansfont[BoldFont={Adagio_Sans}]{Adagio_Sans Light}
\allsectionsfont{\normalfont\sffamily\bfseries}

@@

\begin{abstract}
    While RNA-Seq has enabled great progress towards the goal of wide-scale
    isoform-level mRNA quantification, short reads have limitations when
    resolving complex or similar sets of isoforms. As a result, estimates of
    isoform abundance carry far more uncertainty than those made at the gene
    level. When confronted with this uncertainty, commonly used methods produce
    estimates that are unstable; small perturbations in the data often produce
    dramatically different results confounding downstream analysis. We introduce
    a new method, Isolator, which analyzes all samples in an experiment in
    unison using a simple Bayesian hierarchical model. Combined with aggressive
    bias correction, it produces estimates that are simultaneously accurate and
    stable. In a comprehensive comparison of accuracy and stability, we show
    that this property is unique to Isolator. We further demonstrate that the
    approach of modeling an entire experiment enables new analyses by examining
    splicing monotonicity across several time points in the development of human
    cardiomyocyte cells.
\end{abstract}

@@

Since its introduction, RNA-Seq has rapidly become a preferred method of
studying gene expression, having proved to be reliable, reproducible, and
increasingly affordable. In principle, RNA-Seq enables measurements of
expression at a scale and resolution than has previously been impossible or
impractical. Rather than relying on an often vaguely defined notion of a gene,
expression of specific isoforms or exons can be measured. In practice this
promise is difficult to realize.  There are no standardized names or symbols for
isoforms, annotations are often incomplete or in disagreement, and methods
designed to detect changes in splicing rarely concur \cite{Liu:2014dn}.

Most RNA-Seq procotocols produce short reads (typically 30 to 150 nucleotides).
Because a short read is often compatible with many isoforms, the signal produced
from sequencing must be deconvoluted, requiring a certain level of algorithmic
sophistication. This problem was tackled early on and with great success by
Cufflinks \cite{Trapnell:2010kd}, and indeed many of the methods developed in
the interim have followed from the same basic model of RNA-Seq in which
transcripts are represented as particular distributions over possible reads, and
the problem is then to infer mixing coefficients (relative expression values)
inducing a mixture model that best explains the data \cite{Pachter:2011wm}.

@@

An efficient Markov chain Monte Carlo algorithm is used to generate samples over
the parameters of this model, which are saved in an HDF5 \cite{hdf5} based
format. The output from the sampler can then be processed to produce point
estimates, credible intervals, posterior probabilities, and diagnostics. Though
designed to be run on multiple samples concurrently, Isolator can be run as a
conventional tool on individual samples.

The broad effect of this model is to encode the common assumption that
genes tend to be similarly expressed and similarly spliced between
replicates, and to a lesser extent between conditions. In the absence of
sufficient evidence to the contrary, this informative prior shrinks
estimates towards a baseline of no change, producing more conservative
estimates of effect size than would be obtained by using, for example, a
flat Dirichlet prior. This feature of the model is particularly
important for robust treatment of low-expression transcripts. With few reads
there is little power to determine splicing patterns. Considering the
samples in isolation will often produce very different estimates.
Statistical tests based on these estimates, without considering the full
posterior distribution, make approximations by either either
disregarding the relative or compositional nature of the data and
assuming independence, or by relying on asymptotic properties that may
not hold when a gene is not deeply sequenced.

In addition to this hierarchical model, we attempt to aggressively correct for
technical effects. The naive model of RNA-Seq assumes fragments are sampled
uniformly at random from transcripts in proportion to their abundance. Several
have observed that this uniformity assumption is far from an accurate
description of real sequencing data \cite{Hansen:2010if}.  Though severe at the
nucleotide resolution, gene level expression estimates are only moderately
affected by this bias, as it tends to average out over long transcripts. When
one considers expression at the isoform level, attempting to correct for this
nonuniformity becomes vital, as the difference between two isoforms is often
only a few nucleotides.

We previously developed an efficient and versatile model to account for sequence
specific bias that commonly occurs at the ends of fragments \cite{Jones:2012hg}
In Isolator, we supplemented this with methods accounting for 3' bias, fragment
GC-content bias, and positional bias caused by end-effects of fragmentation.

@@

\section{Results}

The convention for evaluating the accuracy of RNA-Seq analysis methods has
consisted chiefly of comparing point estimate accuracy in simulations,
supplemented with comparisons to qPCR or other technologies. Tuning methods
exclusively to maximize accuracy on simulated data risks solving only an
idealized mathematical model of RNA-Seq while disregarding noise, bias, and
nonuniformity. As no gold standard for whole-transcriptome isoform-level
estimation exists, we adopt multi-faceted approach, with a focus on measuring
the stability or consistency of estimates, along with accuracy.

Comparisons of gene or transcript expression are sensitive to how one
chooses to measure similarity or distance. Because gene expression
typically varies across many orders of magnitude, measurements like
the Pearson correlation, $L_1$ or $L_2$ distance, or root-mean-square
error are often dominated by a small subset of the most highly expressed
genes. Recognizing this, a popular alternative is the Spearman rank
correlation, which conversely can overemphasize the majority of
transcripts which have no, or very low expression. These issues are
discussed in depth by Lovell et al \cite{Lovell:2015il}, who propose
measurements of proportionality as a more meaningful metric. We adopt
their ``proportionality correlation'', defined as
$\text{pcor}(x, y) = 2\text{cov}(\log(x), \log(y)) / (\text{var}(\log(x)) + \text{var}(\log(y)))$.
Similar to other measures of correlation, proportionality correlation
varies from -1 to 1, with 1 indicating perfectly proportionality, and -1
perfect reciprocality. Zeros are accounted for by converting all measurements to
transcripts per million (TPM) and adding 0.1.

Many methods have been proposed to aid in the analysis of RNA-Seq
experiments. We limit our discussion specifically to those that seek to
estimate the relative abundance a set of known transcripts. These
include, non-exhaustively, Cufflinks \cite{Trapnell:2010kd},
RSEM \cite{Li:2011cb},
eXpress \cite{Roberts:2012dh},
BitSeq \cite{Glaus:2012ek},
Sailfish \cite{Patro:2014jd},
Salmon (a refinement of Sailfish currently in development), and
Kallisto \cite{Bray:2015uj}.

RSEM was used to produce both maximum likelihood and posterior
mean estimates, which we label ``RSEM/ML'' and ``RSEM/PM'', respectively.
This is not an exhaustive account of such methods, but does represent a
wide variety of popular approaches.

\subsection{Agreement with qPCR and spike-in controls}

@@

We use data generated by the Sequencing Quality Control Consortium (SEQC)
\cite{Consortium:2014kc}, consisting of four reference samples (labeled A, B, C,
and D). The RNA-seq methods were run using approximately 25 million reads per
sample, obtaining estimates that were then compared to qPCR in Tables
\ref{table:taqman} and \ref{table:primepcr}, and the known ERCC spike-in
mixtures in Table \ref{table:ercc}. In each of these we find that Isolator
produces the highest correlation of the methods compared, though sometimes by a
negligible margin. Despite two different technologies, and two different sets of
genes, the qPCR benchmarks agree closely, ranking the methods nearly
identically. Cufflinks, eXpress, Salmon, and RSEM/ML typically perform similarly
to each other, ith BitSeq, Kallisto, Sailfish, and RSEM/PM trailing slightly.
Results are similar with ERCC spike-in controls, the notable exceptions being
Kallisto and Sailfish, which show higher accuracy, and eXpress, which shows
lower accuracy compared to their relative rankings in the qPCr benchmarks.

\begin{table}
\begin{tabular}[c]{@{}lrrrr@{}}
\toprule\addlinespace
Method & A & B & C & D
\\\addlinespace
\midrule
Isolator & \textbf{0.903} & \textbf{0.908} & \textbf{0.890} &
\textbf{0.884}
\\\addlinespace
eXpress & 0.901 & 0.902 & 0.884 & 0.874
\\\addlinespace
Cufflinks & 0.900 & 0.901 & 0.865 & 0.872
\\\addlinespace
Salmon & 0.899 & 0.897 & 0.882 & 0.871
\\\addlinespace
RSEM/ML & 0.899 & 0.896 & 0.881 & 0.870
\\\addlinespace
Kallisto & 0.893 & 0.885 & 0.874 & 0.860
\\\addlinespace
BitSeq & 0.875 & 0.865 & 0.863 & 0.848
\\\addlinespace
Sailfish & 0.879 & 0.857 & 0.854 & 0.834
\\\addlinespace
RSEM/PM & 0.878 & 0.866 & 0.864 & 0.848
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between gene level quantification
of 806 genes using TaqMan qPCR and RNA-Seq quantification.}
\label{table:taqman}
\end{table}

\begin{table}\begin{tabular}[c]{@{}lrrrr@{}}
\toprule\addlinespace
Method & A & B & C & D
\\\addlinespace
\midrule
Isolator & \textbf{0.878} & \textbf{0.866} & \textbf{0.839} &
\textbf{0.852}
\\\addlinespace
Cufflinks & 0.870 & 0.856 & 0.799 & 0.841
\\\addlinespace
eXpress & 0.870 & 0.855 & 0.829 & 0.840
\\\addlinespace
Salmon & 0.866 & 0.852 & 0.826 & 0.836
\\\addlinespace
RSEM/ML & 0.865 & 0.851 & 0.825 & 0.835
\\\addlinespace
BitSeq & 0.840 & 0.821 & 0.802 & 0.813
\\\addlinespace
Kallisto & 0.858 & 0.840 & 0.817 & 0.826
\\\addlinespace
Sailfish & 0.844 & 0.814 & 0.797 & 0.802
\\\addlinespace
RSEM/PM & 0.840 & 0.822 & 0.803 & 0.811
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between gene level quantification
of 18353 genes using PrimePCR qPCR and RNA-Seq quantification.}
\label{table:primepcr}
\end{table}

\begin{table}\begin{tabular}[c]{@{}lrrrr@{}}
\toprule\addlinespace
Method & A & B & C & D
\\\addlinespace
\midrule
Isolator & \textbf{0.979} & \textbf{0.978} & \textbf{0.981} &
\textbf{0.982}
\\\addlinespace
Salmon & 0.976 & 0.975 & 0.978 & 0.979
\\\addlinespace
Kallisto & 0.972 & 0.972 & 0.973 & 0.976
\\\addlinespace
Sailfish & 0.970 & 0.969 & 0.969 & 0.972
\\\addlinespace
Cufflinks & 0.967 & 0.969 & 0.972 & 0.974
\\\addlinespace
RSEM/PM & 0.943 & 0.949 & 0.944 & 0.949
\\\addlinespace
RSEM/ML & 0.941 & 0.948 & 0.945 & 0.951
\\\addlinespace
BitSeq & 0.940 & 0.949 & 0.943 & 0.946
\\\addlinespace
eXpress & 0.931 & 0.939 & 0.935 & 0.942
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between known proportions of 92
ERCC spike-in controls and RNA-Seq quantification.}
\label{table:ercc}
\end{table}

@@

The four SEQC samples consist of two commercially available reference RNA
samples, labeled $A$ and $B$, and two synthetic samples formed by mixing $A$ and
$B$ in specific proportions. Sample $C$ is composed of 75\% $A$ and 25\% $B$,
while sample $D$ is 25\% $A$ and 75\% $B$. This design allows us to directly
measure expression consistency. If we estimate transcript abundance for each
sample, producing estimates $a, b, c, d$, then for these estimates to be in
agreement we should expect $c$ to be approximated by $0.25a + 0.75b$, and
similarly $d$ by $0.75a + 0.25b$. To capture the notion of conistency, we
measured the closeness of this approximation using proportionality correlation
(Table \ref{table:seqctranscripts}).

@@

Estimates made at the gene level (Supplementary Table \ref{table:seqcgenes})
agree with these results, but are of uniformly higher correlation and show a
much smaller gap between posterior mean and maximum likelihood estimate.
Methods reporting posterior mean estimates (Isolator, BitSeq, RSEM/PM) again
show the highest correlation and are in very close agreement.

\begin{table}\begin{tabular}[c]{@{}lrr@{}}
\toprule\addlinespace
Method & $c$ vs $0.75a + 0.25b$ & $d$ vs $0.25a + 0.75b$
\\\addlinespace
\midrule
Isolator & \textbf{0.975} & \textbf{0.975}
\\\addlinespace
BitSeq & 0.967 & 0.967
\\\addlinespace
RSEM/PM & 0.968 & 0.967
\\\addlinespace
Sailfish & 0.932 & 0.925
\\\addlinespace
RSEM/ML & 0.922 & 0.919
\\\addlinespace
Salmon & 0.916 & 0.914
\\\addlinespace
Kallisto & 0.907 & 0.902
\\\addlinespace
eXpress & 0.903 & 0.899
\\\addlinespace
Cufflinks & 0.870 & 0.916
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between transcript-level estimates
for the mixed samples C and D and weighted averages of estimates for A
and B, corresponding to the mixture proportions for C and D.}
\label{table:seqctranscripts}
\end{table}

@@

\begin{figure*}
\includegraphics[width=\textwidth]{batch-effect-figure.pdf}
\caption{
\textbf{a} A heatmap showing pairwise proportionality correlation between the
samples sequenced on two flowcells each at five sites. Flowcells are numbered 1
or 2 and sequencing sites are abbreviated with three letter codes: Australian
Genome Research Facility (AGR), Beijing Genome Institute (BGI), Cornell
University (CNL), Mayo Clinic (MAY), and Novartis (NVS). \textbf{b} The absolute
change in correlation induced by enabling bias correction for methods that
support it. For clarity this plot excludes points for BitSeq estimates of "MAY
2", as bias correction has an extreme detrimental effect on these.}
\label{fig:batcheffects}
\end{figure*}

We further examined the question of consistency by comparing the same
samples sequenced on different flow cells and at different sites, again
using data from the SEQC. We used 10 samples from the larger SEQC
dataset, each consisting of a single lane from two separate
flowcells sequenced at five different sites, all on Illumina HiSeq 2000
instruments, and compared pairwise agreement between estimates from these
samples (Figure \ref{fig:batcheffects}a).

These correlations are somewhat smaller than those in Table
\ref{table:seqctranscripts},
largely because of shallower sequencing: there five flowcell lanes were
combined, totaling \textasciitilde{}25 millions reads per sample, here
only one lane or \textasciitilde{}5 million reads are used.
Nevertheless, we again see posterior mean estimates with significantly higher
agreement between pairs.

When accounting for batch effects, bias correction methods can have a
significant impact. To examine the efficacy of bias correction, we disabled this
functionality on those methods that support it, repeated the experiment, and
measured the change in pairwise correlation (Figure \ref{fig:batcheffects}b). We
see that bias correction is largely beneficial for each method. Isolator,
Salmon, and Cufflinks show similar improvements with bias correction, though in
a few cases Cufflinks slightly decreases agreement. Kallisto show a consistent,
but very slight improvement. BitSeq's bias correction sometimes has a very positive
effect, but other times negative or disastrous effect. Estimates from the ``MAY
2'' samples in particular had far worse agreement with other samples with bias
correction enabled. Here bias correction has dramatic and detrimental effects on
some number of transcripts, driving the correlation to negative numbers.

While this benchmark is informative, it should be considered a lower bound on
the batch effects and bias: these samples were identically prepared and
sequenced. The fact that technical effects were so clearly present even under
such controlled circumstances is indicative that RNA-Seq is not so reproducible
that batch effects can be, under any circumstanced ignored.

@@

To more directly demonstrate Isolator's capacity to produce more accurate
estimates by modeling entire experiments, we simulated a simple RNA-Seq
experiment consisting of two conditions each with three replicated. Expression
values were generated from a two-component log-normal model with parameters fit
to Cufflinks estimates of data from Kuppusamy et al \cite{Kuppusamy:2015ey}. We
generated 10 million, 100nt paired-end simulated RNA-Seq reads per sample using
rlsim \cite{Sipos:2013tz}.

In this benchmark, Isolator significantly improves on existing methods.
It has the unique advantage of sharing information between replicates.
With the assumption that genes tend to be similarly spliced between
replicates, we are able to more effectively resolve transcript
expression in complex loci. Kallisto, Salmon, RSEM/ML, and Cufflinks all
perform very similarly. While comparisons to qPCR showed eXpress and
Cufflinks with a significant advantage, effective bias correction is
less important in simulated data, plausibly explaining the difference.
Although rlsim models some forms of bias, technical effects in RNA-Seq
are not entirely understood, so these are likely understated in its model.

\begin{table}\begin{tabular}[c]{@{}ll@{}}
\toprule\addlinespace
Method & Correlation
\\\addlinespace
\midrule
Isolator & 0.919
\\\addlinespace
Kallisto & 0.887
\\\addlinespace
Salmon & 0.886
\\\addlinespace
RSEM/ML & 0.881
\\\addlinespace
Cufflinks & 0.881
\\\addlinespace
eXpress & 0.825
\\\addlinespace
Sailfish & 0.816
\\\addlinespace
RSEM/PM & 0.806
\\\addlinespace
BitSeq & 0.796
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between ground truth and estimates
produced by each method on simulated RNA-Seq.}
\label{table:rlsim}
\end{table}

@@

RNA-Seq reads become more informative with length and quantity. We
compared data from the same sample sequenced twice, once with 300nt
paired-end reads on a MiSeq sequencer and again with 100nt paired-end
reads using a HiSeq 2000, yielding approximately 4.5 million and 11
million reads, respectively. To elucidate the effect of deeper
sequencing, we sampled without replacement smaller subsets of the 11
million HiSeq 2000 reads and measured correlation between estimates made using
these subsamples with estimates from the MiSeq data (Figure \ref{fig:rhemac}).
